---
title: "Block Bootstrap"
date: "2021-09-10"
author: "Kittipos Sirivongrungson"
output:
  html_document:
    df_print: paged
    code_folding: "show"
    toc: TRUE
    toc_float: TRUE
    code_download: TRUE
---

```{r setup, include=TRUE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file()) # Set WD to Root
here::i_am("lab_mod/ch5-bootstrap.Rmd")
library(tidyverse)
library(here)
theme_set(theme_bw())
```

I will demonstrate how to do block bootstrap using **Xy** dataset from ISLR's practice questions in the chapter 5.

## Load Data

```{r load}
Xy <- get(load(here("data/5.R.RData")))
```


# Let's Explore Data

Data Frame `Xy` has 3 numeric variables 

```{r}
str(Xy)
```

```{r}
summary(Xy)
```

# Fit Linear Model

Supposed we want to fit a linear regression model of `y` on `X1` and `X2`. 

### Q1: What is the standard error for coefficients of X2 or ($\beta_2$) ?

```{r Xy_lm.fit}
Xy_lm.fit <- lm(y ~ X1 + X2, data = Xy)

summary(Xy_lm.fit)
```

```{r, include=FALSE}
X2_std.err <- sqrt(diag(vcov(Xy_lm.fit)))["X2"]
```

As you can see that $S.E.$ of $\beta_2$ was `r X2_std.err`

### Helper function 

Write a helper function to shuffles rows of `Xy`, fits the Linear model and return $\beta_2$.

```{r Xy_lm_boot_fn}
Xy_lm_boot_fn <- function(data, index){
  
  lm_coef <- coef(lm(y ~ X1 + X2, data = data, subset = index))
  lm_coef["X2"]
  
}
```

Let's test the some iterations of the bootstapped $\beta_2$.

```{r}
set.seed(123)

index1 <- sample(1000, 1000, replace = TRUE)
index2 <- sample(1000, 1000, replace = TRUE)

Xy_lm_boot_fn(Xy, index1)
Xy_lm_boot_fn(Xy, index2)
```

OK, It works!

# Regular Bootstrap

I will use `{boot}` package to compute bootstrap of 1000 replicates using `Xy_lm_boot_fn()` as `statistic` argument in `boot()`.

```{r}
library(boot)
```

```{r}
Xy_lm_boot <- boot(Xy, Xy_lm_boot_fn, R = 1000)

Xy_lm_boot
```

The $S.E.$ of the bootstapped $\beta_2$ is `r sd(Xy_lm_boot$t)`

```{r eval=FALSE, include=FALSE}
sd(Xy_lm_boot$t)
```


# Auto-Correlation

**Why we need to do block bootstrap ?** 

Let's see plots of observation in x-axis and each variables in y-axis.
I will use `{ggplot2}` to do this. 

First I need to add rownames to column and pivot to long format.

```{r Xy_to_long}
Xy_to_long <- function(data) {
  
  data %>% 
    tibble::rownames_to_column("Observations") %>% 
    dplyr::mutate(Observations = as.numeric(Observations)) %>% 
    tidyr::pivot_longer(cols = X1:y, names_to = "Variables", values_to = "Values") %>% 
    dplyr::mutate(Variables = factor(Variables, levels = c( "y", "X1", "X2")))
    
}
```


```{r Xy_long}
Xy_long <- Xy_to_long(Xy)
head(Xy_long)
```

And then plot using `geom_line`:

```{r obs_plot1}
Xy_long %>%   
  ggplot(aes(Observations, Values, color = Variables, linetype = Variables)) +
  geom_line(aes(group = Variables)) +
  labs(title = "Auto-Correlation between Variables", 
       caption = "\"Xy\" dataset from ISLR")
```

(You can also do the same thing by using base R plot by `matplot(Xy,type="l")` as in ISLR's example.)

From this plot, It can be seen that there are correlation between observations (rows).
A given data point (e.g. $(X1_i, X2_i, Y_i)$) would influence the value of next data points (e.g. $(X1_{i+1}, X2_{i+1}, Y_{i+1})$). This is called auto-correlation.

Thus, using regular bootstap might rearrange the consecutive rows that are correlated and modify the result.

**Let's plot observation vs values after bootstapped.**

```{r}
set.seed(123)

## Shuffle Rows
index1 <- sample(1000, 1000, replace = TRUE)
Xy_row_shuffle1 <- Xy[index1, ]


rownames(Xy_row_shuffle1) <- 1:1000
```


```{r obs_plot2}
Xy_to_long(Xy_row_shuffle1) %>%   
  ggplot(aes(Observations, Values, color = Variables, linetype = Variables)) +
  geom_line(aes(group = Variables)) +
  labs(title = "Lost Auto-Correlation after Bootstrapped", 
       caption = "bootstapped \"Xy\" dataset from ISLR")
```


# Block Bootstrap

### Helper function

I've created a helper function to obtain an indexes of block bootstrap with following arguments.

-   `n_block`: Total number of blocks
-   `size`: Number of observation in each block

```{r get_block_index}
get_block_index <- function(n_block, # Number of block
                            size # Number of observation in each block
                            ) {
  
  block1 <- 1:size
  
  step <- 0:(n_block-1)*size
  step_boot <- sample(step, n_block, replace = TRUE)
    
  step_expand <- rep(step_boot, each = size)
  block_expand <- rep(block1, n_block)
  
  block_expand + step_expand

}
```

Let's test `get_block_index()` function. 

For example: 3 blocks with 5 observations each.

```{r}
set.seed(123)

get_block_index(n_block =  3,size =  5)
get_block_index(n_block =  3,size =  5)
```

Ok, Looks like it works!

### Create Block Bootstrap Dataset

I will add "block" column containing numeric vector form 1-10 to flag each block so we can keep tract of data when we performed a block bootstrap.

```{r Xy_block10}
Xy_block10 <- Xy
# Add "block" column to keep tract
Xy_block10$block <- factor(rep(1:10, each = 100))

summary(Xy_block10)
```

Now, try simulate block bootstapped dataset:

```{r}
set.seed(123)
index_block10_a <- get_block_index(n_block = 10, size = 100)
index_block10_b <- get_block_index(n_block = 10, size = 100)

Xy_block10[index_block10_a, ]$block %>% unique()
Xy_block10[index_block10_b, ]$block %>% unique()
```

## Block bootstrapped of Linear Model

Finally, I will perform block bootstrapped of the coefficient $\beta_2$.

Recall that $\beta_2$ is a coefficient of `X2` in the linear regression model of `y` on `X1` and `X2`.

I will need to create this final wrapper function `Xy_lm_boot_fn_block()`.
The main argument were the followings:

-   `data`: Xy dataset
-   `n_block`: Total number of blocks to bootstrap
-   `size`: Number of observation in each block 
-   `R`: The number of bootstrap replicates


```{r Xy_lm_boot_fn_block}
Xy_lm_boot_fn_block <- function(data,
                                n_block = 10, size = 100, 
                                R,
                                return_raw = FALSE
                                ) {
  
  coeffs <- numeric(n_block)
  for (i in 1:R) {
    
    index <- get_block_index(n_block, size)
    coeffs[i] <- Xy_lm_boot_fn(data = data, index = index)
    
  }

  if(return_raw) return(coeffs)
  list(original = mean(coeffs),
       std.err = sd(coeffs))
  
}
```


## Final Results

Block bootstrapped of linear model `y` on `X1` and `X2`; then extract mean and standard error of bootstrapped coefficient `X2` ($\beta_2$).

-   Number of block = 10
-   Size of each block = 100
-   Replicate = 1000

```{r}
Xy_lm_boot_fn_block(Xy_block10, n_block = 10, size = 100, R = 1000)
```

Compare it with regular bootstrap

```{r}
Xy_lm_boot
```

And with no bootstrap (`X2`)

```{r}
summary(Xy_lm.fit)
```

### Plot of bootstrapped coefficients

```{r Xy_boot_coeffs}
Xy_boot_coeffs <- Xy_lm_boot_fn_block(Xy_block10, n_block = 10, size = 100, R = 1000, 
                    return_raw = TRUE)
```


```{r hist, message=FALSE}
library(latex2exp)

data.frame(beta2 = Xy_boot_coeffs) %>% 
  ggplot(aes(beta2)) +
  geom_histogram(fill = "grey", color = "black") +
  
  geom_vline(aes(xintercept = mean(Xy_boot_coeffs)), color = "red") +
  annotate("label", x = mean(Xy_boot_coeffs), y = 125, 
           label = "mean", 
           color = "red") +
  
  labs(x = TeX("$\\hat{\\beta^*_2}$"), y = "Count", 
       title = TeX("Distribution of block bootstrapped $\\beta_2$")) 
```



